---
title: "Dowtown Commuter Market"
author: "Raven McKnight"
date: "9/13/2019"
output:
    MetroTransitr::metro_html: default
    MetroTransitr::metro_report: default
---
<style type="text/css">
html {
  max-width: 10.5in;
  margin: 1.25in 0.75in;
  font-family: "Calibri", helvetica, sans-serif;
  font-size: 11pt;
}

body { width: 8.5in;
       max-width: 10.5in;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(data.table)
library(ggplot2)
library(tigris)
library(sf)
library(leaflet)
library(dplyr)
library(htmltools)
library(rgdal)

tot <- readRDS('data/tot_jobs.RDS')
ods <- readRDS('data/od_jobs.RDS')

tot_ys <- tot[, .(workers = sum(workers)), by = year]

options(tigris_class = 'sf')
options(tigris_use_cache = FALSE)
counties <- c('Anoka', 'Carver', 'Dakota', 'Hennepin', 'Ramsey', 'Scott', 'Washington')
bgs <- block_groups('MN', counties, year = 2016)

bgs <- st_transform(bgs, 4326)

# downtown area
dt <- st_bbox(c(xmin = -93.286765, ymin = 44.965247, xmax = -93.25244, ymax = 44.98825))
dt <- st_as_sfc(dt, crs = 4326)
st_crs(dt) <- 4326
dt <- st_sf(dt)

# this is so messy but best i have right now
dtbg <- st_intersection(bgs, dt)
dtbgs <- unique(dtbg$GEOID)
dtbg <- bgs %>% filter(GEOID %in% dtbgs)
```

# Total downtown jobs

This map shows total jobs in the downtown Minneapolis area in 2017. Colors are logged, labels show unlogged values. 
```{r}
# total jobs in 2017 ------------------
tot_17 <- tot[year == 2017][, .(workers = sum(workers)), by = w_bg]

bg_tot_17 <- left_join(dtbg, tot_17, by = c('GEOID' = 'w_bg'))

emp2017pal <- colorNumeric(palette = 'YlGnBu', domain = log(bg_tot_17$workers))
emp2017labs <- sprintf(
  "<strong> %s </strong> jobs in 2017",
  bg_tot_17$workers
) %>% lapply(htmltools::HTML)

leaflet(bg_tot_17) %>% addTiles() %>%
  addPolygons(weight = 1, color = 'grey', fillColor = ~emp2017pal(log(workers)), 
              fillOpacity = 0.7, label = emp2017labs)
```


# Change in downtown jobs
This map shows the *change* in jobs from 2010-2017 in the same extent. 
```{r}
# change 2010-2017
tot_10 <- tot[year == 2010][, .(workers10 = sum(workers)), by = w_bg]
change_10_17 <- tot_10[tot_17, on = 'w_bg']
change_10_17[, change := workers - workers10]
change_10_17[, logchange := sign(change)*log(abs(change))]

change_10_17 <- left_join(dtbg, change_10_17, by = c('GEOID' = 'w_bg'))

change1017pal <- colorNumeric(palette = 'RdBu', domain = c(-10.1, 10.1))
change1017labs <- sprintf(
  "Block group ID: %s <br> Change in jobs 2010-2017: <strong> %s </strong>",
  change_10_17$GEOID, change_10_17$change
) %>% lapply(htmltools::HTML)

leaflet(change_10_17) %>% addTiles() %>%
  addPolygons(weight = 1, color = 'grey', fillColor = ~change1017pal(logchange),
              fillOpacity = 0.7, label = change1017labs)
```

# Downtown employee origins

This map shows the origins of employees who work in the region above in 2017. There are lots of block groups much further away with 1-3 employees listed but I'm guessing those people are, for the most part, not actually commuting daily. I included existing park & ride lots as well. 
```{r message = FALSE, warning = FALSE, results = "hide"}
dtw <- ods[year == 2017 & w_bg %in% dtbgs]
dtw[, dtworkers := sum(workers), by = 'h_bg']

# maybe overboard but i want context for the p&rs
empcounties <- c('Sherburne', 'Wright', 'Chisago', 'Isanti', 'Le Sueur', 'Mille Lacs', 
                 'Sibley', 'Stearns', 'Meeker', 'Benton', 'McLeod', 'Morrison', 
                 'Todd', 'Kanabec', 'Pine', 'Nicollet', 'Rice', 'Goodhue')
wicounties <- c('St. Croix', 'Pierce', 'Pepin', 'Eau Clair', 'Dunn', 'Chippewa', 
                'Polk', 'Barron', 'Burnett', 'Buffalo', 'Trempealeau')

empbg <- block_groups('MN', county = empcounties, year = 2016)
wibg <- block_groups('WI', county = wicounties, year = 2016)

empbg <- st_transform(empbg, 4326)
wibg <- st_transform(wibg, 4326)

allbg <- rbind_tigris(empbg, wibg, bgs)
  
dtw_origins <- left_join(allbg, dtw, by = c('GEOID' = 'h_bg'))

# p&r -----------------------
prurl <- 'ftp://ftp.gisdata.mn.gov/pub/gdrs/data/pub/us_mn_state_metc/trans_park_and_ride_lots/shp_trans_park_and_ride_lots.zip'
loc <- file.path(tempdir(), 'pr.zip')
download.file(prurl, loc)
unzip(loc, exdir = file.path(tempdir(), 'pr'), overwrite = TRUE)
file.remove(loc)
pr <- readOGR(file.path(tempdir(), 'pr'), layer = 'ParkAndRideLots', stringsAsFactors = FALSE)
pr <- st_as_sf(pr)
pr <- st_transform(pr, 4326)

# active only
pr <- pr %>% filter(Active == 1)

originpal <- colorNumeric(palette = 'YlGnBu', domain = c(0, 7), na.color = 'transparent')
originlab <- sprintf(
  "<strong> %s </strong> downtown workers in 2017",
  dtw_origins$dtworkers
) %>% lapply(htmltools::HTML)

prlab <- sprintf(
  "<strong> %s </strong> <br> %s",
  pr$Name, pr$PrPlRec
) %>% lapply(htmltools::HTML)
```
```{r}
leaflet(dtw_origins) %>% addTiles() %>%
  addPolygons(weight = 1, color = 'grey', fillColor = ~originpal(log(dtworkers)),
              fillOpacity = 0.7, label = originlab) %>%
  addCircles(data = pr, label = prlab, color = 'purple')
```


# Change in downtown employee origins
Finally, this map shows the change in the origins above from 2010-2017. 
```{r}
dtw10 <- ods[year == 2010 & w_bg %in% dtbgs]
dtw10[, dtworkers10 := sum(workers), by = 'h_bg']
dtw10 <- unique(dtw10[, c('dtworkers10', 'h_bg')])
dtw <- unique(dtw[, c('dtworkers', 'h_bg')])

dtwc <- dtw10[dtw, on = 'h_bg']
dtwc[, change := dtworkers - dtworkers10]

dtw_change <- left_join(allbg, dtwc, by = c('GEOID' = 'h_bg'))

changepal <- colorNumeric(palette = 'RdBu', domain = c(-6.5, 6.5), na.color = 'transparent')
changelab <- sprintf(
  "Change in downtown workers 2010-2017: <strong> %s </strong>",
  dtw_change$change
) %>% lapply(htmltools::HTML)

leaflet(dtw_change) %>% addTiles() %>%
  addPolygons(weight = 1, color = 'grey', fillColor = ~changepal(sign(change) * log(abs(change))),
              fillOpacity = 0.7, label = changelab) %>%
  addCircles(data = pr, label = prlab, color = 'purple', radius = 1)
```


# Next steps

* Filtering to isolate more specific origin block groups - ie, places that aren't already served? Places where new/expanded P&R may be necessary?
* Put O-D data into an app (like the Streetlight O-D app)?